ranacapa: An R package to explore environmental DNA data with exploratory statistics and interactive visualizations

### 2018-10-02 manuscript.Rmd
updated draft of the ms; figs

The targeted amplification and sequencing of DNA that living organisms shed into the physical environment they occupy (termed “environmental DNA metabarcoding”, or “eDNA sequencing”)  is revolutionizing microbiology, ecology, and conservation research [@Taberlet_2012; @Deiner_2017].  Sequencing of environmental DNA extracted from field-collected soil, water, or sediment samples holds great promise to shed light on a range of questions, ranging from tracking the dynamics of bacterial communities or profiling the composition of ancient plant and animal communities [@Props_2016; @Pedersen_2014], to motoring populations of rare or endangered species [@Balasingham_2017]. As the cost of eDNA sequencing declines and sample collection techniques become more streamlined (e.g. @Thomas_2018), professional research scientists are increasingly using eDNA sequencing as a platform to partner with members of the community, such as natural resource managers, undergraduate students, and citizen scientists (collectively referred to in this manuscript as “community scientists”), in primary research. However, developing robust and impactful community science programs that engage community partners in all steps of the research process remains a challenge.

eDNA sequencing-based projects work well for community science partnerships because non-experts can be quickly trained to collect samples in the field, and because eDNA sequencing is an exciting framework for research pertinent to disciplines such as medicine, agriculture, ecology, and geography [@Deiner_2017]. Community partners in such programs can have heterogeneous backgrounds, ranging from curious members of the public for whom collecting samples in the field is the first scientific research experience (e.g. University of California’s CALeDNA program, http://www.ucedna.com/), to professional natural resource managers who regularly collaborate with research scientists (e.g. Center for Ocean Solutions’ eDNA project, https://oceansolutions.stanford.edu/project-environmental-dna). In these partnerships (as in any other), community participants should be able to participate across multiple stages of the project, not only in sample collection [@Pandya_2012; @ECSA_2015]. This can be a challenge for eDNA sequencing-based community science programs because although it is relatively easy to train community partners to collect eDNA samples, it is far more challenging to train them to interact with and visualize the results from these studies: learning the bioinformatic tools necessary for visualizing and analyzing the large, multidimensional datasets generated in eDNA sequencing studies can be difficult for professional researchers (Carey & Papin 2018), let alone for the diverse community partners of community science programs.

To address this challenge, we created an R package “ranacapa”, at the core of which is a Shiny webapp with which users can visualize and perform simple community ecology analyses on results from eDNA sequencing studies. `ranacapa` complements existing visualization platforms (e.g. Phinch, Phyloseq-Shiny, QIIME2 Viewer), because in addition to interactive visualizations, `ranacapa` includes brief explanations and links to additional educational resources to provide users with an overview of basic data analyses used in eDNA studies. `ranacapa` works with community matrices generated via QIIME () and stored as BIOM tables or with community matrices generated with the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), which is used extensively by the CALeDNA program.
In the remainder of this manuscript, we describe `ranacapa` and demonstrate its use by two community science partnerships based at the University of California, Los Angeles (UCLA): first, a collaboration between eDNA researchers and resource managers at the National Park Service, and second, a partnership between community ecology researchers and an undergraduate microbiology course at UCLA. As we show in the Use Cases, empowering community partners to interact with the data and perform simple but insightful community ecology analyses can help make these collaborations more enriching and valuable to both parties.

## Implementation
At the core of `ranacapa` is a Shiny webapp (@Chang_2018), and the package also includes two categories of helper functions (Table 1). The first set of functions works to connect the taxonomy tables, generated either by the Anacapa eDNA sequence analysis pipeline (https://github.com/limey-bean/Anacapa; Curd et al. in prep) or QIIME (Caporaso 2010), into Phyloseq objects that can be used for downstream visualizations and analyses. The second set of functions, which includes two externally written functions openly available on GitHub, extends the visualization and statistical functionality of the `phyloseq` [@McMurdie_2013] and `vegan` (@Oksanen_2018) packages.

The Shiny app (http://gauravsk.shinyapps.io/ranacapa or `rancapa::runRanacapaApp()`) allows users to interact with eDNA results through statistical summaries and interactive plots, displayed in the following tabs:

- **Sample rarefying**: This tab explains the basic logic behind rarefying samples in metagenomics studies, and allows users to choose the sampling depth (Figure 1). Users can control the number of times the samples are rarified. The documentation on this tab also acknowledges recent disagreement regarding the value of rarefying in metabarcoding and eDNA sequencing studies (@McMurdie_2014), and allows users to proceed through the rest of the app without rarefying samples.

@@

- **Taxonomy heatmap**: This tab shows the taxonomy-by-sample matrix as an interactive heatmap made using `heatmaply::heatmaply()`, where the color of each cell represents the number of times a given taxon was sequenced in a sample (Figure 2).

@@

- **Taxonomy barplot**: This tab shows the taxonomy-by-sample matrix as an interactive barplot (Figure 3).

@@

- **Alpha diversity plots**: This tab introduces the concept of Alpha diversity as the local diversity measured in a single habitat or sample.  Users can plot Alpha diversity as observed taxon richness or as Shannon Diversity per sample, or can group samples according to a variable in the metadata file (Figure 4).

@@

- **Alpha diversity statistics**: This tab allows users to choose a variable from the metadata, and generates an Alpha diversity ANOVA table according to the user-selected variable. The tab also shows the output from  a post-hoc Tukey test.

- **Beta diversity plots**: This tab introduces the concept of Beta diversity as the turnover in species composition across habitats (or samples). The tab includes an ordination plot generated by `phyloseq::plot_ordination()` based on an ordination made with `phyloseq::ordinate(., method = "PCoA")`. Points on the PCoA plot are colored according to a user-selected metadata variable (Figure 5). The tab also shows dissimilarity among sites according in a dendrogram generated with Ward’s cluster analysis (`stats::hclust(distance_object, method = "ward.d2")`, where `distance_object` is made using `phyloseq::distance()`. For both figures, users can toggle between using Jaccard and Bray-Curtis dissimilarity.

@@

- **Beta diversity statistics**: This tab shows results from two statistical tests of species turnover across site: first, a multivariate ANOVA implemented with `vegan::adonis())`, which shows users results from a statistical test of taxon turnover across sites, and second, a statistical test of heterogeneity of variances among samples implemented with `vegan::betadisper()`, which shows results from a statistical test that compares the degree of sample-to-sample variation within habitats (or within other user-selected groups).

- **More references**

## Operation
`ranacapa` depends on Bioconductor v 3.7, which in turn relies on `R` v 3.5.0. The Shiny app has been tested on Chrome and Firefox on Windows, Mac-OSX, and Ubuntu. The package can be installed using the command `devtools::install_github("gauravsk/ranacapa")`, and the Shiny app is available at http://gauravsk.shinyapps.io/ranacapa.

### Input file structure
`ranacapa` generates interactive visualizations based on two input files. The first required input is a site-by-species matrix, uploaded either as a `.biom` or as a tab-separated `.txt` file. If the site-by-species matrix is uploaded as a `.biom` file, the file should contain taxonomy and abundance information; thus, `.qza` files generated by QIIME2 are not immediately suitable for ranacapa. If the site-by-species matrix is uploaded as a `.txt` file, the file should match the specifications of the output files from the Anacapa eDNA sequence analysis pipeline (). Specifically, each row in the `.txt` file must represent the taxonomic identification for one Amplicon Sequence Variant (ASV), and each column, save one, must represent the number of times that ASV appears in each sequenced sample. One column, named `sum.taxonomy` must contain the taxonomic identification, with taxonomic rank separated by a `;`. as follows:

@@

The second required input is a tab-separated `.txt` file that contains sample metadata. The first column in this metadata file should match the column names in the sample names in the taxonomy table; the remaining columns of the metadata file contain sample information for each of the samples in the site-by-species matrix A valid metadata file for the taxonomy table above would be as follows:

@@

The `ranacapa` function `validate_input_files()` verifies that both the taxonomy table and the metadata files match certain structural requirements, which are documented in the function help files. The current version of `ranacapa` accepts both categorical and continuous metadata columns, but in the latter case, continuous values are categorized into bins.

## Use Cases
We designed `ranacapa` to be used by eDNA researchers to share the results from their research with community partners. Specifically, we expect that researchers with bioinformatic expertise will use best-practices to assign taxonomy to eDNA datasets using the pipeline of their choice and generate clean taxonomy and metadata files. Researchers will then use `ranacapa` to share results with their community partners, emphasizing the analyses or visualizations most appropriate to their use case. We document two such partnerships below that showcase how `ranacapa` can facilitate authentic communication between researchers and community scientists.

@@

eDNA research scientists can use `ranacapa` to share results, especially interactive taxonomy lists, with natural resource managers. For example `ranacapa` was used by researchers at UCLA who partner with resource managers at the Channel Islands National Park to assess the potential for eDNA as a biodiversity monitoring tool in the Southern California Channel Islands. The goal of this ongoing collaboration is to assess whether eDNA metabarcoding studies can provide insights to supplement ongoing management efforts at the park, which are currently done with expensive and time-intensive visual surveys (@Lessios_1996, @Murphy_2010, @Usseglio_2015). Implementing streamlined eDNA-based monitoring may allow a dramatic expansion in the scope and scale of marine ecosystem assessment in the California (@Edgar_2007, @Deiner_2017).

To begin exploring whether eDNA-based studies can supplement visual underwater surveys, resource managers at the Channel Islands National Park Service collected and filtered thirty-1L water samples for eDNA analysis at permanent monitoring sites inside and adjacent to MPAs in the park. Research scientists at UCLA performed metabarcode sequencing of the mitochondrial 12S and CO1 gene regions from these samples targeting bony fishes, elasmobranches, and invertebrate taxa. The researchers processed sequences and assigned taxonomy using the Anacapa toolkit. When taxonomy tables were ready,  researchers used the `ranacapa` Shiny app to share results from this pilot study with National Park resource managers.

The taxonomy heatmap (Figure 3) was the most valuable vizualation to this collaboration, because it allowed the resource managers to focus on a particular set of key taxa. The heatmap showed that this pilot study detected 36 of the 70 key metazoan taxa monitored by the managers at the species level, and the remaining 34 at the genus, family, or order level. This indicates that eDNA-based studies can likely supplement ongoing management efforts and provide new insights into the spatial and temporal distributions of these key species, especially rare and difficult to observe taxa such as endangered or invasive species. The resource managers were also interested in the PCoA plot, which was used to explore whether well-known major biogeographic patterns in the Channel Islands (e.g. turnover of fish communities across gradients in sea surface temperature, ) are detected using eDNA analyses. The value of `ranacapa` in this scenario was to highlight the strengths and areas for concern in using eDNA to monitor diversity in the Channel Islands. Due to the potential for eDNA to help improve detection of rare species (especially endangered species or newly introduced exotics), which are difficult to observe visually, this collaboration is continuing. The data from this study are packaged as the demo dataset for the `ranacapa` Shiny web-app and are available online at XXX.

@@

Students can use `ranacapa` to interact with results from metabarcoding studies and to learn the basic structure of eDNA datasets. A research-based environmental microbiology course at UCLA [@Shapiro_2015] used eDNA metabarcoding approaches to study the impact of a recent local wildfire on the plant and soil microbial community. The goal of this twenty-week course was to provide students an authentic experience in basic microbiology and microbial community ecology research. The instructors helped students develop a research question, design a sampling regime to test their hypotheses, and conduct fieldwork to collect soil samples for eDNA analyses in burned and unburned natural areas. Over the first ten weeks of the course, the instructional team (which included eDNA research scientists) extracted total DNA and sequenced the ITS2 [@Gu_2013] and 16S SSU RNA [@Caporaso_2012] metabarcoding region to characterize the plant, bacterial, and archaeal communities in the student-collected soil samples. The researchers then processed the sequences and assigned taxonomy using the Anacapa toolkit.

Shortly after taxonomy tables were generated, the course instructors introduced students to eDNA data exploration and simple statistical analyses using ranacapa. A key strength of using `ranacapa` was that despite having no prior bioinformatics experience, students began exploring on the an online instance of Shiny app (http://gauravsk.shinyapps.io/ranacapa) within a single class period. Thus, using `ranacapa` allowed the instructors to focus their time with the students on biological questions rather than on troubleshooting bioinformatics problems, as had been the case in previous sessions of the course. The course instructors noted that this basic exploration in ranacapa, which was not part of the curriculum in previous iterations of the course, had several positive impacts on students and their research projects. First, `ranacapa` helped students explore the basic structure of the dataset and begin to understand the relationships between community profile and the various metadata they had collected in the field. Second, `ranacapa` opened the door to basic diversity analyses– for example, students could easily test their hypotheses regarding the taxonomic diversity of microbes in burned and unburned soils.  Third, by significantly reducing the time and difficulty in visualizing soil microbial diversity patterns, `ranacapa` helped students develop and pursue more sophisticated analyses during the remaineder of the course using tools such as STAMP. The taxonomy tables and metadata files used in this course are available online at XXXX.

@@

We propose three avenues for future work in ranacapa. First, we plan to use `ranacapa` as the primary tool to present eDNA results from hundreds of samples sequenced by the CALeDNA community science program. The positive experience with reserve managers suggests open forums to discuss `ranacapa` output will be fruitful to strengthen the feedback loop between community partners and researchers. Second, `ranacapa` will be a key tool in the upcoming undergraduate curriculum module “Pipeline for Undergraduate Microbiome Analysis”, which is being built as a complete suite of analysis and data visualization tools which will be made openly available to undergraduate researchers. Finally, in the long-term, we believe that there is great promise in connecting `ranacapa` to packages that connect with APIs of online biodiversity databases (e.g. Taxize, rinat). This will help users explore a much wider range of biodiversity questions, for example, by programmatically asking whether their samples include invasive species that are absent from other nearby sites. Such apps that allow non-technical audiences to easily interact with results from eDNA sequencing studies have great potential to engage community partners with a wide range of backgrounds and interests in primary research.

@@

## Grant information
GSK and ZJG were supported by the US-NSF Graduate Research Fellowship (DEG No. 1650604) during the development of this package. EEC is supported by the CALeDNA program, with funds from University of California President’s Research Catalyst Award (CA-16-376437).

## Acknowledgments
We thank Sabrina Shirazi, Rachel Turba, Chris Dao, and Keith Mitchell for providing feedback on developmental versions of this package. `ranacapa` builds on numerous functions that have been made openly available online with a GPL-3 License, namely the “phyloseq-extended” toolkit written by Mahendra Mariadassau (https://github.com/mahendra-mariadassou/phyloseq-extended) and “pairwise.adonis” written by Pedro Martinez Arbizu (https://github.com/pmartinezarbizu/pairwiseAdonis).

>>>>

The targeted amplification and sequencing of DNA that living organisms shed into their physical environment, termed "environmental DNA metabarcoding" or "eDNA sequencing",  is revolutionizing microbiology, ecology, and conservation research [@Taberlet_2012; @Deiner_2017].  Sequencing of environmental DNA extracted from field-collected soil, water, or sediment samples can shed light on a range of questions, ranging from profiling the composition of ancient plant and animal communities [@Pedersen_2014], to motoring populations of rare or endangered species [@Balasingham_2017]. As the cost of eDNA sequencing declines and sample collection techniques become more streamlined (e.g. @Thomas_2018), professional research scientists are increasingly using eDNA sequencing as a platform to partner with a diversity of community partners, including natural resource managers, undergraduate students, and citizen scientists in primary research. However, developing robust and impactful community science programs that engage community partners in all steps of the research process remains a challenge.

eDNA sequencing-based projects work well for community science partnerships because non-experts can be quickly trained to collect samples in the field, and because eDNA sequencing is an exciting framework for research pertinent to disciplines such as medicine, agriculture, ecology, and geography [@Deiner_2017]. Community partners in such programs can have heterogeneous backgrounds, ranging from curious members of the public for whom collecting samples in the field is the first scientific research experience (e.g. University of California’s CALeDNA program, http://www.ucedna.com/), to professional natural resource managers who regularly collaborate with research scientists (e.g. Center for Ocean Solutions’ eDNA project, https://oceansolutions.stanford.edu/project-environmental-dna). A key ingredient to ensure sustained engagement of community partners is that they should be able to participate across multiple stages of the research project, not only in sample collection [@Pandya_2012; @ECSA_2015]. This can be a challenge for eDNA sequencing-based community science programs because although it is relatively easy to train community partners to collect eDNA samples, it is far more challenging to train them to interact with and visualize the results from these studies. Indeed, learning the bioinformatic tools necessary for visualizing and analyzing the large, multidimensional datasets generated in these studies can be difficult for professional researchers (Carey & Papin 2018), let alone for the non-technical audience of community science programs.

To address this challenge, we created an R package “ranacapa”, at the core of which is a Shiny webapp with which users can visualize and perform simple community ecology analyses on results from eDNA sequencing studies. `ranacapa` complements existing visualization platforms (e.g. Phinch, Phyloseq-Shiny, QIIME2 Viewer), because in addition to interactive visualizations, `ranacapa` includes brief explanations and links to additional educational resources to provide users with an overview of basic data analyses used in eDNA studies. `ranacapa` works with community matrices generated via QIIME () or the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), which is used extensively by the CALeDNA program.

In the remainder of this manuscript, we describe `ranacapa` and demonstrate its use by two community science partnerships based at the University of California, Los Angeles (UCLA): first, a collaboration between eDNA researchers and resource managers at the National Park Service, and second, a partnership between community ecology researchers and an undergraduate microbiology course at UCLA. As we show in the Use Cases, empowering community partners to interact with the data and perform simple but insightful community ecology analyses can help make these collaborations more enriching and valuable to both parties.

## Implementation
At the core of `ranacapa` is a Shiny webapp (@Chang_2018), which is avialable at http://gauravsk.shinyapps.io/ranacapa or with `ranacapa::runRanacapa()`. The package also includes two categories of helper functions (Table 1) that transform user-uploaded taxonomy and metadata tables into `R` objects that can be visualized and analuzed using Phyloseq [@McMurdie_2013] and Vegan [@Oksanen_2018]. `ranacapa` is available for installation on Github or CRAN:

```{r eval = F, echo = T}
devtools::install_github("gauravsk/ranacapa")
install.packages("ranacapa")
```

The `ranacapa` Shiny app allows users to interact with eDNA results through statistical summaries and interactive plots, displayed in the following tabs:

- **Sequencing depth**:  Introduces the potential for variation in sequencing depth among samples and explains the basic logic behind rarefying samples in metagenomics studies. Allows users to rarefy the dataset to a sampling depth (Figure 1). The documentation on this tab also acknowledges recent disagreement regarding the value of rarefying in metabarcoding and eDNA sequencing studies [@McMurdie_2014], and allows users to proceed through the rest of the app without rarefying samples.

@@

- **Taxonomy heatmap**: Shows the taxonomy matrix as an interactive heatmap made using `heatmaply::heatmaply()`, where the color of each cell represents the number of times a given taxon was sequenced in a sample (Figure 2).

@@

- **Taxonomy barplot**: Shows the taxonomy-by-sample matrix as an interactive barplot (Figure 3).

@@

- **Alpha diversity plots**: Introduces the concept of Alpha diversity as the local diversity measured in a single habitat or sample. Users can plot Alpha diversity as observed taxon richness or as Shannon Diversity per sample, or can group samples according to a variable in the metadata file (Figure 4).

@@

- **Alpha diversity statistics**: Allows users to choose a variable from the metadata, and generates an Alpha diversity ANOVA table according to the user-selected variable. The tab also shows the output from  a post-hoc Tukey test.

- **Beta diversity plots**: Introduces the concept of Beta diversity as the turnover in species composition across habitats (or samples). The tab includes an ordination plot generated by `phyloseq::plot_ordination()` based on an ordination made with `phyloseq::ordinate(., method = "PCoA")`. Points on the PCoA plot are colored according to a user-selected metadata variable (Figure 5). The tab also shows dissimilarity among sites according in a dendrogram generated with Ward’s cluster analysis (`stats::hclust(distance_object, method = "ward.d2")`), where `distance_object` is made using `phyloseq::distance()`. For both figures, users can toggle between using Jaccard and Bray-Curtis dissimilarity.

@@

- **Beta diversity statistics**: Shows results from two statistical tests of species turnover across site: first, a multivariate ANOVA implemented with `vegan::adonis())`, which shows users results from a statistical test of taxon turnover across sites, and second, a statistical test of heterogeneity of variances among samples implemented with `vegan::betadisper()`, which shows results from a statistical test that compares the degree of sample-to-sample variation within habitats (or within other user-selected groups).

## Operation
`ranacapa` depends on Bioconductor v 3.7, which in turn relies on `R` v 3.5.0. The Shiny app has been tested on Chrome and Firefox on Windows, Mac-OSX, and Ubuntu.

### Input file structure
The `ranacapa` Shiny app requires two input files. The first requirement is a site-by-species matrix, uploaded either as a rich, dense `.biom` table, or as a tab-separated `.txt` file. Qiime2-generated`.qza` files generated by QIIME2 are not immediately suitable for ranacapa, as they do not contain full taxonomy information. If the site-by-species matrix is uploaded as a `.txt` file, the file should match the specifications of the output files from the Anacapa eDNA sequence analysis pipeline (). Specifically, each row in the `.txt` file must represent a taxonomic identification , and each column (save one) must represent the number of times that ASV appears in each sequenced sample. One column, named `sum.taxonomy` must contain the taxonomic identification, with taxonomic rank separated by a semicolon, as follows:

@@

The second requirement is a tab-separated `.txt` file that contains sample metadata. The first column should match the sample names in the taxonomy table; the remaining columns of the metadata file contain sample information for each of the samples in the site-by-species matrix. The metadata should contain categorical variables with two or more categories per variable. A valid metadata file for the taxonomy table above would be as follows:

@@

The `ranacapa` function `validate_input_files()` verifies that both the taxonomy table and the metadata files match structural requirements, which are documented in the function help files.

## Use Cases
We expect that researchers with bioinformatic expertise will use the pipeline of their choice to assign taxonomy to eDNA datasets and generate clean taxonomy and metadata files that can be visualized in `ranacapa`. Researchers can share these files with their partners, and emphasize the analyses or visualizations most appropriate to their use case. We now show how `ranacapa` can facilitate authentic communication between researchers and community partners in two settings.

@@

We expect the interactive taxonomy heatmap to be especially useful for researchers collaborating with natural resource managers whose efforts are focused a targetted list of rare or invasive taxa. For example, UCLA researchers who partner with resource managers at the Channel Islands National Park Service to assess the potential for eDNA as a biodiversity monitoring tool to supplement expensive and time-intensive visual biodiversity surveys in the Southern California Channel Islands [@Lessios_1996, @Usseglio_2015; @Deiner_2017] use ranacapa to share eDNA sequencing results. For this partnership, resource managers collected and filtered thirty-1L water samples for eDNA analysis at permanent monitoring sites inside and adjacent to protected areas, and research scientists at UCLA performed eDNA sequencing of the mitochondrial 12S and CO1 gene, targeting bony fishes, elasmobranches, and invertebrate taxa. The researchers processed sequences and assigned taxonomy using the Anacapa toolkit, and shared results with the resource managers using the `ranacapa` Shiny app.

The taxonomy heatmap (Figure 2) was the most valuable visualization to this collaboration, because it allowed the resource managers to filter the large observed species list down to a particular set of key taxa that they regularly monitor. The heatmap showed that this pilot study detected 36 of the 70 key metazoans at the species level, and the remaining 34 at the genus, family, or order level. This indicates that eDNA-based studies can likely supplement ongoing management efforts and provide new insights into the spatial and temporal distributions of these species. The value of `ranacapa` in this scenario was to quickly sort through long species lists generated in by eDNA sequencing to highlight the strengths and weaknesses in using eDNA to monitor diversity in the Channel Islands. The data from this study are packaged as the demo dataset for the `ranacapa` Shiny web-app and are available online at XXX.

@@

We expect `ranacapa`'s exploratory visualizations and the accompanying explanations to help researchers introducing ecology concepts and analyses in classrooms or other informal education settings. For example, `ranacapa` was used by a research-based environmental microbiology course at UCLA [@Shapiro_2015], in which students used eDNA metabarcoding to study the impact of a local wildfire on the plant and soil microbial community. The goal of this twenty-week course was to provide undergraduate students an authentic experience in basic microbiology and microbial community ecology research. Over the first ten weeks, the instructional team, which included eDNA research scientists, extracted total DNA from student-collected soil samples and sequenced the ITS2 [@Gu_2013] and 16S SSU RNA [@Caporaso_2012] metabarcoding region to characterize the plant, bacterial, and archaeal communities. The instructional team processed sequences and assigned taxonomy using the Anacapa toolkit.

Students were introduced to the structure of eDNA sequencing results and were taught to explore data and perform simple statistical analyses using `ranacapa`. A key benefit of using `ranacapa` was that despite having no prior bioinformatics experience, students could begin exploring the biodiversity in their samples in a matter of minutes by using the online instance of Shiny app. This allowed the instructors to focus classroom time on biological questions rather than on troubleshooting bioinformatics problems, as had been the case in previous sessions of the course. The course instructors noted that visualizing eDNA data in `ranacapa` helped students understand the relationships between community profile and the various metadata they had collected in the field. By significantly reducing the time and difficulty in visualizing basic biodiversity patterns, `ranacapa` helped students develop and pursue more sophisticated analyses during the remainder of the course, using most sophisticated tools such as STAMP [@Parks_2014] and PICRUSt [@Langille_2013]. The taxonomy tables and metadata files used in this course are available online at XXXX.

@@

We propose three avenues for future work in ranacapa. First, we plan to use `ranacapa` as the primary tool to present eDNA results from hundreds of samples sequenced by the CALeDNA community science program.  Second, `ranacapa` will be a key tool in the upcoming undergraduate curriculum module "Pipeline for Undergraduate Microbiome Analysis", which is being built as an open-source, comprehensive suite of analysis and data visualization tools for undergraduate researchers. Finally, in the long-term, we believe there is great promise in connecting `ranacapa` to packages that connect with APIs of online biodiversity databases (e.g. Taxize, rinat). This will help users explore a much wider range of biodiversity questions, for example, by programmatically asking whether their samples include invasive species that are absent from other nearby sites. In sum, tools like `ranacapa` that allow non-technical audiences to easily interact with results from eDNA sequencing studies have great potential to engage community partners with a wide range of backgrounds and interests in primary research.

@@

## Grant information
GSK and ZJG were supported by the US-NSF Graduate Research Fellowship (DEG No. 1650604). EEC is supported by the CALeDNA program, with funds from University of California President’s Research Catalyst Award (CA-16-376437).

## Acknowledgments
We thank Sabrina Shirazi, Rachel Turba, Chris Dao, and Keith Mitchell for providing feedback on developmental versions of this package. We also thank Mahendra Mariadassau and  Pedro Martinez Arbizu for making their `phyloseq-extended` (https://github.com/mahendra-mariadassou/phyloseq-extended)  and `pairwiseAdonis` (https://github.com/pmartinezarbizu/pairwiseAdonis) packages openly available with a GPL-3 License.
